---
title: 'Assignment 5B: Chess Elo Calculations'
author: "Ummay Rukiya"
date: "2026-10-02"
output: html_document
---

## Overview

For this assignment, I will use the chess tournament data from Project 1 to compare each player’s actual score with the score predicted by the Elo rating system. I will calculate the expected result of every game using the rating difference between each player and their opponent. After adding each player’s expected results, I will identify the five players who performed furthest above expectations and the five who performed furthest below expectations.

## Loading the Tournament Data

I loaded the original tournament text file directly from my Project 1 GitHub repository. Using the raw GitHub file makes the analysis reproducible without requiring a file stored on my computer.

```{r load-data}
data_url <- paste0(
  "https://raw.githubusercontent.com/UR-71/",
  "DATA607-Project1/main/tournamentinfo.txt"
)

tournament_lines <- readLines(
  data_url,
  warn = FALSE
)

length(tournament_lines)
```
## Extracting Players and Opponents

Each player’s tournament record appears across two lines. I identified the player lines and rating lines, separated the fields at the vertical bars, and extracted each player’s number, name, actual score, pre-tournament rating, and opponents.

```{r extract-records}
player_lines <- tournament_lines[
  grep("^\\s*[0-9]+\\s*\\|", tournament_lines)
]

rating_lines <- tournament_lines[
  grep("^\\s*[A-Z]{2}\\s*\\|", tournament_lines)
]

split_record <- function(record) {
  trimws(strsplit(record, "|", fixed = TRUE)[[1]])
}

player_fields <- lapply(player_lines, split_record)
rating_fields <- lapply(rating_lines, split_record)

pair_number <- as.integer(
  sapply(player_fields, function(record) record[1])
)

player_name <- sapply(
  player_fields,
  function(record) record[2]
)

actual_score <- as.numeric(
  sapply(player_fields, function(record) record[3])
)

rating_text <- sapply(
  rating_fields,
  function(record) record[2]
)

pre_rating <- as.integer(
  sub(".*R:\\s*([0-9]+).*", "\\1", rating_text)
)

round_results <- lapply(
  player_fields,
  function(record) record[4:10]
)

extract_opponents <- function(rounds) {
  completed_games <- grepl(
    "^[WLD]\\s*[0-9]+",
    rounds
  )

  opponent_ids <- sub(
    "^[WLD]\\s*([0-9]+).*$",
    "\\1",
    rounds[completed_games]
  )

  as.integer(opponent_ids)
}

opponent_numbers <- lapply(
  round_results,
  extract_opponents
)

length(player_name)
length(opponent_numbers)
```

## Calculating Expected Scores

I used the Elo expected-score formula shown in the reference video. For each completed game, I compared the player’s pre-tournament rating with the opponent’s pre-tournament rating. I then added the individual game expectations to calculate each player’s total expected score.

The formula is:

**Expected score = 1 / (1 + 10^((opponent rating - player rating) / 400))**

Source: singingbanana. (2019, February 15). *The Elo Rating System for Chess and Beyond* [Video]. YouTube. https://www.youtube.com/watch?v=AsYfbmp0To0

```{r calculate-expected-scores}
rating_lookup <- setNames(
  pre_rating,
  pair_number
)

calculate_expected_score <- function(
    player_rating,
    opponent_ids
) {
  opponent_ratings <- rating_lookup[
    as.character(opponent_ids)
  ]

  game_expectations <- 1 / (
    1 + 10^(
      (opponent_ratings - player_rating) / 400
    )
  )

  sum(game_expectations)
}

expected_score <- mapply(
  calculate_expected_score,
  player_rating = pre_rating,
  opponent_ids = opponent_numbers
)

performance_difference <- actual_score - expected_score

elo_results <- data.frame(
  pair_number,
  player_name,
  pre_rating,
  actual_score,
  expected_score = round(expected_score, 2),
  performance_difference = round(
    performance_difference,
    2
  ),
  stringsAsFactors = FALSE
)

head(elo_results)
```

## Five Most Overperforming Players

A positive performance difference means that a player earned more tournament points than the Elo formula predicted. I arranged the results from the largest positive difference to the smallest and selected the first five players.

```{r top-overperformers}
overperformers <- elo_results[
  order(
    elo_results$performance_difference,
    decreasing = TRUE
  ),
]

top_five_overperformers <- head(
  overperformers,
  5
)

knitr::kable(
  top_five_overperformers,
  caption = "Five Players Who Most Overperformed"
)
```

## Five Most Underperforming Players

A negative performance difference means that a player earned fewer tournament points than the Elo formula predicted. I arranged the results from the most negative difference upward and selected the first five players.

```{r top-underperformers}
underperformers <- elo_results[
  order(elo_results$performance_difference),
]

top_five_underperformers <- head(
  underperformers,
  5
)

knitr::kable(
  top_five_underperformers,
  caption = "Five Players Who Most Underperformed"
)
```
## Interpretation

Aditya Bajaj overperformed the most. Based on the ratings of his opponents, he was expected to earn approximately 1.95 points, but he earned 6 points. His actual score was therefore about 4.05 points higher than expected. The other strongest overperformers were Zachary James Houghton, Anvit Rao, Jacob Alexander Lavalley, and Amiyatosh Pwnanandam.

Loren Schwiebert underperformed the most. He was expected to earn approximately 6.28 points but earned 3.5 points, producing a difference of -2.78. George Avery Jones, Jared Ge, Rishi Shetty, and Joshua David Lee were the other four players who performed furthest below their expected scores.

## Validating the Results

I checked that the completed results contained all 64 players and that none of the required calculation columns contained missing values.

```{r validate-results}
nrow(elo_results)

colSums(
  is.na(
    elo_results[
      c(
        "pre_rating",
        "actual_score",
        "expected_score",
        "performance_difference"
      )
    ]
  )
)
```

## Conclusion

The Elo calculations showed that tournament results did not always match what the players’ ratings predicted. Some lower-rated players performed much better than expected against their opponents, while some higher-rated players earned fewer points than expected.

Aditya Bajaj had the largest positive difference between his actual and expected scores, while Loren Schwiebert had the largest negative difference. This demonstrates how the Elo formula can provide a useful baseline for identifying unusually strong or weak tournament performances.



























